---
title: "Exploratory Data Analysis"
author: "Nathan W. Van Bibber"
date: "9/10/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

> "EDA is a state of mind... not a formal process with a strict set of rules"

## Data cleaning

Do the data meet your expectations?

Need all the tools of EDA:

* visualization
* transformation
* modeling


```{r}
library(tidyverse)
library(nycflights13)
library(GGally)
```


The `GGally::ggpairs()` function displays a graphic of all the variables plotted against each other. 
```{r}
GGally::ggpairs(diamonds)

## or filter down to fewer variables
diamonds %>% 
    select(cut, carat) %>% 
    ggpairs()
```


Note can put the `aes()` on a line by itself too.
```{r}
ggplot(diamonds) +
    aes(x = cut, y = price) +
    geom_jitter(alpha = 0.05, color = "blue") +
    geom_boxplot()
    
```

```{r}
diamonds %>% 
    count(color, cut) %>% 
    ggplot() +
    geom_tile(aes(x = color, y = cut, fill = n)) +
    coord_equal()
```

Compare the volume, weight, and purity of the diamonds.

```{r}
diamonds_purity <- 
    diamonds %>% 
    mutate(volume = x * y * z,
           weight = carat * 0.2, # weight in grams: 1 carat ~ 0.2 grams
           purity = (weight/ volume)/ 0.00352 # pure diamond ~ 3520 kg/m^3
    )
 diamonds_purity %>% 
     select(volume, weight, purity) %>% 
     ggpairs()


diamonds_purity %>% 
    filter(purity > 1, is.finite(purity))

ggplot(diamonds_purity) +
    aes(x = weight, y = volume) +
    geom_point() +
    geom_abline(slope = 1/0.00352, intercept = 0)
    geom_smooth()

ggplot(diamonds_purity) +
    aes(x = purity) + 
    geom_histogram()
```

