---
title: "Data Transformation: Notes and Exercises"
author: "Nathan W. Van Bibber"
date: "1/17/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Transformation

```{r me, message=FALSE}
library(nycflights13)
library(tidyverse)
```

### Filter rows with `filter()`

allows you to subset observations based on their values. 

* The first argument is the *name* of the data frame. 
* The second and subsequent arguments are the *expressions* that filter the data frame


### Common problems with floating point numbers:

```{r eval=FALSE}
sqrt(2) ^ 2 == 2
#> [1] FALSE
1 / 49 * 49 == 1
#> [1] FALSE
```

Computers use finite precision arithmetic (can’t store an infinite # of digits!) so remember that

* **every number you see is an approximation**. 
* Instead of relying on `==`, use `near()`

#### The `%in%` operator

A useful short-hand is `x %in% y`. This will select every row where x is one of the values in y

#### De Morgan’s law

* !(x & y) is the same as !x | !y
* !(x | y) is the same as !x & !y

#### Tip

Whenever you start using complicated, multipart expressions in filter(), consider making them explicit variables instead. That makes it much easier to check your work.

#### Missing values

```{r eval=FALSE}
NA > 5
#> [1] NA
10 == NA
#> [1] NA
NA + 10
#> [1] NA
NA / 2
#> [1] NA
NA == NA
#> [1] NA
```


`filter()` only includes rows where the condition is TRUE; it excludes both FALSE and NA values. If you want to preserve missing values, ask for them explicitly

#### Exercises

1. Find all flights that:
    - Had an arrival delay of two or more hours
    - Flew to Houston (IAH or HOU)
    - Were operated by United, American, or Delta
    - Departed in summer (July, August, and September)
    - Arrived more than two hours late, but didn’t leave late
    - Were delayed by at least an hour, but made up over 30 minutes in flight
    - Departed between midnight and 6am (inclusive)  
  
2. Another useful dplyr filtering helper is between(). What does it do? Can you use it to simplify the code needed to answer the previous challenges?

3. How many flights have a missing dep_time? What other variables are missing? What might these rows represent?

4. Why is NA ^ 0 not missing? Why is NA | TRUE not missing? Why is FALSE & NA not missing? Can you figure out the general rule? (NA * 0 is a tricky counterexample!)

### Arrange rows with `arrange()`

similar to filter() except that instead of selecting rows, it changes their order. 
It takes: 
* a data frame and 
* a set of column names (or more complicated expressions) to order by. 
    - If more than one column name, each additional column will be used to break ties in the values of preceding columns
* Use `desc()` to re-order by a column in descending order
* Missing values are always sorted at the end

#### Exercises

1. How could you use arrange() to sort all missing values to the start? (Hint: use is.na()).

2. Sort flights to find the most delayed flights. Find the flights that left earliest.

3. Sort flights to find the fastest (highest speed) flights.

4. Which flights traveled the farthest? Which traveled the shortest?


### Select columns with `select()`

allows you to rapidly zoom in on a useful subset using operations based on the names of the variables.
* It can also rename selected varaibles

#### Helper functions

* `everything()`: useful if you have a handful of variables you’d like to move to the start of the data frame.

* `starts_with("abc")`: matches names that begin with “abc”.

* `ends_with("xyz")`: matches names that end with “xyz”.

* `contains("ijk")`: matches names that contain “ijk”.

* `matches("(.)\\1")`: selects variables that match a regular expression. This one matches any variables that contain repeated characters. You’ll learn more about regular expressions in strings.

* `num_range("x", 1:3)`: matches x1, x2 and x3.

See `?select` for more details

#### Exercises

1. Brainstorm as many ways as possible to select dep_time, dep_delay, arr_time, and arr_delay from flights.

2. What happens if you include the name of a variable multiple times in a select() call?

3. What does the `any_of()` function do? Why might it be helpful in conjunction with this vector?
```{r}
vars <- c("year", "month", "day", "dep_delay", "arr_delay")
```

4. Does the result of running the following code surprise you? How do the select helpers deal with case by default? How can you change that default?

```{r}
select(flights, contains("TIME"))
```


### Add new variables with `mutate()`

always adds new columns at the end of your dataset 
    
If you only want to keep the new variables, use `transmute()`

#### Useful creation funcitons

must be vectorised: it must take a vector of values as input, return a vector with the same number of values as output. 

* Arithmetic operators: `+`, `-`, `*`, `/`, `^`.
* Modular arithmetic: `%/%` (integer division) and `%%` (remainder)
* Logs: `log()`, `log2()`, `log10()` 
    - All else being equal, I recommend using log2() because it’s easy to interpret: a difference of 1 on the log scale corresponds to doubling on the original scale and a difference of -1 corresponds to halving.
* Offsets: l`ead()` and `lag()` allow you to refer to leading or lagging values
* Cumulative and rolling aggregates: functions for running sums, products, mins and maxes: `cumsum()`, `cumprod()`, `cummin()`, `cummax()`
    - dplyr provides `cummean()` for cumulative means. 
    - If you need rolling aggregates (i.e. a sum computed over a rolling window), try the **RcppRoll** package.
* Logical comparisons, `<`, `<=`, `>`, `>=`, `!=`, and `==`, 
* Ranking: there are a number of ranking functions, but you should start with `min_rank()`
    -  variants include: `row_number()`, `dense_rank()`, `percent_rank()`, `cume_dist()`, `ntile()`

#### Exercises

1. Currently dep_time and sched_dep_time are convenient to look at, but hard to compute with because they’re not really continuous numbers. Convert them to a more convenient representation of number of minutes since midnight.

2. Compare air_time with arr_time - dep_time. What do you expect to see? What do you see? What do you need to do to fix it?

3. Compare dep_time, sched_dep_time, and dep_delay. How would you expect those three numbers to be related?

4. Find the 10 most delayed flights using a ranking function. How do you want to handle ties? Carefully read the documentation for min_rank().

5. What does 1:3 + 1:10 return? Why?

6. What trigonometric functions does R provide?

### Grouped summaries with `summarise()`

Together `group_by()` and `summarise()` provide one of the tools that you’ll use most commonly when working with dplyr: grouped summaries

When you group by multiple variables, each summary peels off one level of the grouping  

Note: Be careful when progressively rolling up summaries: it’s OK for sums and counts, but you need to think about weighting means and variances, and it’s not possible to do it exactly for rank-based statistics like the median. In other words, the sum of groupwise sums is the overall sum, but the median of groupwise medians is not the overall median.  

#### Exercises

1. Brainstorm at least 5 different ways to assess the typical delay characteristics of a group of flights. Consider the following scenarios:

    - A flight is 15 minutes early 50% of the time, and 15 minutes late 50% of the time.
    - A flight is always 10 minutes late.
    - A flight is 30 minutes early 50% of the time, and 30 minutes late 50% of the time.
    - 99% of the time a flight is on time. 1% of the time it’s 2 hours late.
    - Which is more important: arrival delay or departure delay?

2. Come up with another approach that will give you the same output as not_cancelled %>% count(dest) and not_cancelled %>% count(tailnum, wt = distance) (without using count()).

3. Our definition of cancelled flights (is.na(dep_delay) | is.na(arr_delay) ) is slightly suboptimal. Why? Which is the most important column?

4. Look at the number of cancelled flights per day. Is there a pattern? Is the proportion of cancelled flights related to the average delay?

5. Which carrier has the worst delays? Challenge: can you disentangle the effects of bad airports vs. bad carriers? Why/why not? (Hint: think about flights %>% group_by(carrier, dest) %>% summarise(n()))

6. What does the sort argument to count() do. When might you use it?

#### Ungrouping
If you need to remove grouping, and return to operations on ungrouped data, use `ungroup()`
