---
title: "Data Transformation: Notes and Exercises"
author: "Nathan W. Van Bibber"
date: "1/17/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Transformation

```{r me, message=FALSE}
library(nycflights13)
library(tidyverse)
```

### Common problems with floating point numbers:

```{r eval=FALSE}
sqrt(2) ^ 2 == 2
#> [1] FALSE
1 / 49 * 49 == 1
#> [1] FALSE
```

Computers use finite precision arithmetic (can’t store an infinite # of digits!) so remember that

* **every number you see is an approximation**. 
* Instead of relying on `==`, use `near()`

### The `%in%` operator

A useful short-hand is `x %in% y`. This will select every row where x is one of the values in y

### De Morgan’s law

* !(x & y) is the same as !x | !y
* !(x | y) is the same as !x & !y

### Tip

Whenever you start using complicated, multipart expressions in filter(), consider making them explicit variables instead. That makes it much easier to check your work.

### Missing values

```{r}
NA > 5
#> [1] NA
10 == NA
#> [1] NA
NA + 10
#> [1] NA
NA / 2
#> [1] NA
NA == NA
#> [1] NA
```


filter() only includes rows where the condition is TRUE; it excludes both FALSE and NA values. If you want to preserve missing values, ask for them explicitly

### Exercises

1. Find all flights that:
    Had an arrival delay of two or more hours
    Flew to Houston (IAH or HOU)
    Were operated by United, American, or Delta
    Departed in summer (July, August, and September)
    Arrived more than two hours late, but didn’t leave late
    Were delayed by at least an hour, but made up over 30 minutes in flight
    Departed between midnight and 6am (inclusive)

2. Another useful dplyr filtering helper is between(). What does it do? Can you use it to simplify the code needed to answer the previous challenges?

3. How many flights have a missing dep_time? What other variables are missing? What might these rows represent?

4. Why is NA ^ 0 not missing? Why is NA | TRUE not missing? Why is FALSE & NA not missing? Can you figure out the general rule? (NA * 0 is a tricky counterexample!)