---
title: "Tibbles"
author: "Nathan W. Van Bibber"
date: "1/29/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Tibbles

some differences from `data.frame()`

* never converts strings to factors
* never changes names of variables
* never creates row names
* column names can be invalid varaible names
    - need backticks `` surrounding 
* With base R data frames, `[` sometimes returns a data frame, and sometimes returns a vector. With tibbles, `[` always returns another tibble.
    - one of the main reasons an older function might not work with a tibble
    - `dplyr::filter()` and `dplyr::select()` allow you to solve the same problems with clearer code

```{r message=FALSE}
library(tidyverse)

tb <- tibble(
  `:)` = "smile", 
  ` ` = "space",
  `2000` = "number"
)
tb
```


Use `as_tibble()` to convert a data frame to a tibble

Use `tribble()` for data entry in code. It makes it possible to lay out small amounts of data in easy to read form.

* column headings are defined by formulas (i.e. they start with `~`), and 
* entries are separated by commas. This 
* its useful to add a comment line, to make it really clear where the header is.

```{r}
tribble(
  ~x, ~y, ~z,
  #--|--|----
  "a", 2, 3.6,
  "b", 1, 8.5
)
```


Explicitly `print()` the data frame and control the number of rows (`n`) and the `width` (number of characters ) of the display

Use `problems()` to see the rows causing issues when trying to create a tibble.